---
title: "Correlational Analysis in R"
author: "Clay Ford"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Email from Jemima

I was wondering if you knew of a way to analyze correlational data with treatments in R? I've written the code for without treatments but I'm unsure of how to make R analyze it via different treatments. The treatments (3) are in the column "Salt"

## Read in data 

```{r message=FALSE}
library(readxl)
library(ggplot2)
library(psych)

licor <- read_excel("Licor Data for R.xlsx")
```

Subset data for variables of interest and include treatment variable

```{r}
d <- licor[,c("A","Ca","Ci","gsw","Salt")]
summary(d) # check for missings
table(d$Salt) # sample sizes within treatment groups
```

## Visualize with scatterplots and trend lines

Get all combinations of two variables

```{r}
var_pairs <- combn(x = c("A","Ca","Ci","gsw"), m = 2)
var_pairs
```

Write a function to help us make the plots for all three levels of Salt.

```{r}
f <- function(df, x, y){
  ggplot(df) +
    aes(x = .data[[x]], y = .data[[y]]) +
    geom_point() +
    geom_smooth() +
    facet_wrap(~Salt) +
    ggtitle(paste(y, "vs", x))
}
```


Apply the function to the variable pairs.

```{r message=FALSE}
apply(var_pairs, 2, function(x)f(df = d, x = x[1], y = x[2]))
```

It looks like Ci vs A and gsw vs A are the only two pairs that exhibit any correlation, though it's not clear there are any differences between the three levels of Salt.



## Single correlation test 

The code in this section and the next come from the help page for the `r.test()` function in the {psych} package.

Perform a single correlation test for gsw vs A where Salt = control and Salt = high. When calculating correlations, I set `method = "spearman"` since the plots showed some non-linearity.

```{r}
# correlation between A and gsw given Salt == "control"
r1 <- cor.test(~ A + gsw, data = d, subset = Salt == "control", 
               method = "spearman")

# correlation between A and gsw given Salt == "high"
r2 <- cor.test(~ A + gsw, data = d, subset = Salt == "high", 
         method = "spearman")

# view the correlations
cbind("Salt = control" = r1$estimate, "Salt = high" = r2$estimate)
```

Now run test to see if the difference in correlations is plausible using the `r.test()` function. 

```{r}
r.test(n = sum(d$Salt == "control"), r12 = r1$estimate, r34 = r2$estimate, 
       n2 = sum(d$Salt == "high"))
```

The probability of seeing a difference this big (or bigger) between the correlations, assuming there is no real difference between the two, is pretty high (p = 0.57). We fail to conclude there is any important difference between the two correlations. 

## Compare two correlation matrices

Compare all pairs of variables where Salt = control and Salt = high. This is probably a little more efficient than doing two at a time. 

First we create the two correlation matrices and view the correlations. To do this I use the `lowerUpper()` function from the {psych} package. The R1 correlations are displayed below the diagonal, while the R2 correlations are displayed above the diagonal.

```{r}
# correlation matrix given Salt == "control"
R1 <- cor(d[d$Salt == "control", -5])

# correlation matrix given Salt == "high"
R2 <- cor(d[d$Salt == "high", -5])

# show the two correlations 
round(lowerUpper(lower = R1, upper = R2), 2)
```

For example, we see the correlation between Ci vs A is -0.63 when salt = control (lower corner) and -0.74 when salt = high (upper corner)

Now run the test to compare all correlations at once.

```{r}
# compare the correlation matrices
test <- r.test(n=sum(d$Salt == "control"), r12 = R1, r34 = R2, 
       n2 = sum(d$Salt == "high"))

# view the p-values, rounded to 3 decimal places
round(test$p, 3)
```

The difference in correlations between Ci vs A may be of interest (p = 0.03). However, given that we ran six different hypothesis tests we may want to adjust the p-values to protect against false positives.

Below we extract the six p-values and use `p.adjust()` to adjust the p-values using the default Holm method. We then insert the adjusted p-values into the upper corner of the p-value matrix, but leave the original p-values in the lower corner.

```{r}
# show the p values of the difference between the two matrices
adjusted <- p.adjust(test$p[upper.tri(test$p)])
both <- test$p
both[upper.tri(both)] <- adjusted

# The lower off diagonal are the raw p-values, 
# the upper off diagonal are the adjusted p-values
round(both,digits=2)
```

After adjustment, the p-value comparing the correlations for Ci vs A is now 0.18. 
The code above can be re-run to compare Salt = "control" to Salt = "low", or Salt = "low" to Salt = "high". Judging from the plots, it seems unlikely any important differences between correlations will be detected. 

## References

- R Core Team (2024). _R: A Language and Environment for Statistical Computing_. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.
- Revelle, W. (2024). _psych: Procedures for Psychological, Psychometric, and Personality Research_. Northwestern University, Evanston, Illinois. R package version 2.4.12, <https://CRAN.R-project.org/package=psych>.
- Wickham, H. (2016) _ggplot2: Elegant Graphics for Data Analysis_. Springer-Verlag New York.
