---
title: "Example of Measurement Invariance"
author: "Clay Ford"
date: "2025-06-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This example comes from Chapter 4 of Beaujean (2014), Latent Variable Models with Multiple Groups. The example examines if the structure of the WISC-III scale is the same in children with and without manic symptoms. [Source](https://blogs.baylor.edu/rlatentvariable/sample-page/r-syntax/#Chapter_4_Latent_Variable_Models_with_Multiple_Groups)

Load the package we need.

```{r}
library(lavaan)
```


## Read in data

The data is manually entered as covariance matrices. This is not normally how you would enter data in R. Notice we enter two separate covariance matrices: one for children with manic symptoms and those without.

```{r}
# variable names
wisc3.names <- c("Info", "Sim", "Vocab","Comp", 
                 "PicComp", "PicArr", "BlkDsgn", "ObjAsmb")

# covariance for group with manic symptoms
manic.cov <- c(9.364, 7.777, 12.461, 6.422, 8.756, 10.112, 
               5.669, 7.445, 6.797, 8.123, 3.048, 4.922, 
               4.513, 4.116, 6.200, 3.505, 4.880, 4.899, 
               5.178, 5.114, 15.603, 3.690, 5.440, 5.220, 
               3.151, 3.587, 6.219, 11.223, 3.640, 4.641, 
               4.877, 3.568, 3.819, 5.811, 6.501, 9.797)

# lavaan function to create a covariance matrix
manic.cov <- lav_matrix_lower2full(manic.cov)

# means of the eight variables
manic.means <- c(10.09, 12.07, 10.25, 9.96, 10.90, 11.24, 10.30, 10.44)

# label the covariances and means
colnames(manic.cov) <- rownames(manic.cov)  <- wisc3.names
names(manic.means) <- wisc3.names

# preview the covariance matrix
manic.cov
```


Now do the same for the children without manic symptoms.

```{r}
# covariance for group without manic symptoms
norming.cov <- c(9.610, 5.844, 8.410, 6.324, 6.264, 9.000, 
                 4.405, 4.457, 5.046, 8.410, 4.464,  4.547, 
                 4.512, 3.712, 10.240, 3.478, 2.967, 2.970, 
                 2.871, 3.802,  10.890, 5.270, 4.930, 4.080, 
                 3.254, 5.222, 3.590, 11.560, 4.297, 4.594, 
                 4.356, 3.158, 4.963, 3.594, 6.620, 10.890)
norming.cov <- lav_matrix_lower2full(norming.cov) 

# means
norming.means <- c(10.10, 10.30, 9.80, 10.10, 10.10, 10.10, 9.90, 10.20)

# label the covariances and means
colnames(norming.cov) <- rownames(norming.cov)  <- wisc3.names
names(norming.means) <- wisc3.names

# preview the covariance matrix
norming.cov
```

Finally, combine the covariance matrices, sample sizes, and means into single list objects.

```{r}
combined.cov <- list(manic = manic.cov, norming = norming.cov)
combined.n <- list(manic = 81, norming = 200)
combined.means <- list(manic = manic.means, norming = norming.means)
```

Now ready to specify the CFA model.

## CFA Model

The model below says there are two factors: Verbal-Comprehension (VC) and Visual-Spatial (VS). The model hypothesizes these two factors are influencing four variables each. The model also says we want to estimate the covariance between the two factors (last line).

```{r}
wisc3.model <-'
VC =~ Info + Sim + Vocab + Comp 
VS =~ PicComp + PicArr + BlkDsgn + ObjAsmb
VC ~~ VS
'
```

We'll also define a vector of fit indices so we can easily request them after fitting a model.

```{r}
# specify fit indices of interest
fit.indices <- c("chisq", "df", "cfi", "rmsea", "srmr", "mfi")
```

## Configural Invariance

This fits the same model to both groups and allows all parameters to be freely estimated. In other words, we're doing a CFA with each group. Notice in the summary output each group's parameters are different.

```{r}
configural.fit <- cfa(wisc3.model, 
                      sample.cov = combined.cov,
                      sample.nobs = combined.n, 
                      sample.mean = combined.means) 
summary(configural.fit)
```

Fit indices look good.

- CFI = comparative fit index (0.95 or greater)
- RMSEA = root mean square error of approximation (0.06 or lower)
- SRMR = standardized root mean square residual (0.08 or below)
- MFI = McDonald's Fit Index (0.95 or greater)

```{r}
fitMeasures(configural.fit, fit.indices)
```

And the residuals are small. Notice everything is smaller than 2. That's good.

```{r}
residuals(configural.fit, type = "normalized")
```

This is a good fitting model for _each group_. 

## Weak Invariance

Now constrain loadings to be equal between groups. Same code as above with one additional argument: `group.equal = "loadings"`. Notice in the summary output that all the loadings (listed under Latent Variables) are equal in both groups. 

```{r}
weak.fit <- cfa(wisc3.model, 
                sample.cov = combined.cov, 
                sample.nobs = combined.n, 
                sample.mean = combined.means, 
                group.equal = "loadings") 
summary(weak.fit)
```

According to fit measures this is also a good model.

```{r}
fitMeasures(weak.fit, fit.indices)
```

And the residuals look good as well, though they are getting bigger.

```{r}
residuals(weak.fit, type = "normalized")
```
Finally, we could formally compare the models using a Chi-squared difference test (aka, Log-likelihood ratio test). We can do this with the `lavTestLRT()` function. The null hypothesis is no difference between the models. A small p-value provides evidence against this hypothesis and suggests a preference for the more complex model (ie, the model with fewer degrees of freedom). Below there is some evidence against the weak invariance model (p = 0.049), however AIC and BIC metrics suggest otherwise. Lower AIC/BIC values are better and I don't see any reason to reject the weak invariance model.

```{r}
lavTestLRT(configural.fit, weak.fit)
```


This is also a good model. It appears the manifest (observed) variables for each group are influenced in the same way by the two factors.

## Strong Invariance

Now constrain both loadings and intercepts to be equal. Notice the `group.equal` argument now has a vector: `c("loadings", "intercepts")`. Notice in the summary output that all the loadings (Latent Variables) and Intercepts are equal in both groups. 

```{r}
strong.fit <- cfa(wisc3.model, 
                  sample.cov = combined.cov, 
                  sample.nobs = combined.n, 
                  sample.mean = combined.means, 
                  group.equal = c("loadings", "intercepts")) 
summary(strong.fit)
```

The fit measures for this model are not so good.

```{r}
fitMeasures(strong.fit, fit.indices)
```

But the residuals are mostly OK. The only point of strain is the residual for the "Sim" mean (2.45) in the manic group.

```{r}
residuals(strong.fit, type = "normalized")
```

The Chi-squared difference test suggests we reject the strong invariance model for the weak invariance model based on the small p-value. 

```{r}
lavTestLRT(weak.fit, strong.fit)
```

For sake of completeness, let's move to the next model.

## Strict Invariance

Now constrain loadings, intercepts and variances to be equal. Notice the `group.equal` argument has the vector: `c("loadings", "intercepts", "residuals")`. Notice in the summary output that all the loadings (Latent Variables), intercepts and variances are equal in both groups. 


```{r}
strict.fit <- cfa(wisc3.model, 
                  sample.cov = combined.cov, 
                  sample.nobs = combined.n, 
                  sample.mean = combined.means, 
                  group.equal = c("loadings", "intercepts", "residuals"))
summary(strict.fit)
```

Fit measures are not too good. The CFI is too low and RMSEA is too high.

```{r}
fitmeasures(strict.fit, fit.indices)
```

However, the residuals look good for the most part. The only point of strain is the residual for the "PicCmp" variance (-2.829) in the manic group.

```{r}
residuals(strict.fit, type = "normalized")
```

The Chi-squared test favors the strong invariance model based on AIC and p-value, but the BIC actually points to the strict fit. 

```{r}
lavTestLRT(strong.fit, strict.fit)
```



## Partial Strict Invariance

If we want, we can remove constraints for certain parameters using the `group.partial` argument. For example, to allow the variance for PicCmp to be estimated separately between the groups, add the line `group.partial = "PicComp ~~ PicComp"`. In the summary, notice under Variances that "PicComp" gets a separate estimate in each group.

```{r}
strict.fit2 <- cfa(wisc3.model, 
                   sample.cov = combined.cov, 
                   sample.nobs = combined.n,
                   sample.mean = combined.means, 
                   group.equal = c("loadings", "intercepts", "residuals"),
                   group.partial = "PicComp ~~ PicComp")
summary(strict.fit2)
```

The fit measures are not great...

```{r}
fitmeasures(strict.fit2, fit.indices)
```

But the residuals look pretty good.

```{r}
residuals(strict.fit2, type = "normalized")
```

Obviously one has to wonder if we would arrive at the same conclusions using a new sample of data. We don't want to build a model that's too specific to our sample. That's would be overfitting. 

## References

- Beaujean, A. A. (2014) _Latent Variable Modeling Using R_. Routledge.
- Beaujean, A. A., Freeman, M. J., Youngstrom, E., & Carlson, G. (2012). The structure of cognitive abilities in youths with manic symptoms: a factorial invariance study. Assessment, 19(4), 462–471. <https://doi.org/10.1177/1073191111399037>
- R Core Team (2025). _R: A Language and Environment for Statistical  Computing_. R Foundation for Statistical Computing, Vienna, Austria.  <https://www.R-project.org/>.
- Rosseel, Y. (2012). lavaan: An R Package for Structural Equation Modeling. _Journal of Statistical Software_, 48(2), 1-36. <https://doi.org/10.18637/jss.v048.i02>
