---
title: "Mediation analysis and t-test"
author: "Clay Ford"
date: '2022-05-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# generate data
n <- 400
set.seed(999)
grp <- sample(0:1, size = n, replace = TRUE) |> factor()
M <- 1.5 + 0.6*(grp=="1") + rnorm(n, sd = 0.1)
Y <- 1.9 + 0.1*(grp=="1") + 0.5*M + rnorm(n, sd = 0.1)
# TRUE ACME/indirect effect = 0.6 * 0.5 = 0.3
# TRUE ADE = 0.1
# TRUE Total Effect = ACME + ADE = 0.3 + 0.1 = 0.4
d <- data.frame(grp = grp, Y = Y, M = M)

```

## The consultation question

I had a phd candidate in systems engineering ask why their mediation analysis was not agreeing with their t-test. (wut?)

## The t-test

Let's start with the t-test. To keep it simple we call their dependent variable `Y` and the treatment `grp`, which is a 0/1 indicator variable. We see that the mean of Y for grp 1 is higher than the mean of Y for grp 0.

```{r}
aggregate(Y ~ grp, data = d, mean)
```

Is that difference "significant"? They carried out a t-test:

```{r}
t.test(Y ~ grp, data = d)
```

Sure does appear so. The 95% CI is (-0.42, -0.37). We're pretty sure the mean of grp 0 is at least -0.37 less than the mean of grp 1. 

The _difference in means_ can be thought of as the _effect size_. Below we take the difference between grp1 and grp 0. I use the base R pipe to pass the estimate (ie, the two means) into the `diff()` function, which takes the first element of a vector and subtracts it from the next element. Then I use `as.numeric()` to strip off some metadata. 

```{r}
t.test(Y ~ grp, data = d)$estimate |> diff() |> as.numeric()
```

The effect of the treatment (grp = 1) is about 0.396. In other words, if you're in grp 1, we expect your Y value to be about 0.396 higher than it otherwise would have been if you were in grp 0.

## The mediation analysis

The student's advisor suggested there was another variable acting as a _mediator_ and that they should do a mediation analysis. Let's call the mediator variable `M`. The basic idea is that the treatment variable, `grp` is actually affecting `M`, which is then affecting `Y`. This is usually visualized with a diagram. A former StatLab grad student associate wrote a [good blog post](https://data.library.virginia.edu/introduction-to-mediation-analysis/) on this topic and works through an example with some pictures. 

Three quantities of interest usually come out of a mediation analysis:

1. The effect of the mediator, also called the "average causal mediation effect" (ACME) or indirect effect.
2. The effect of treatment _controlling for mediation_, also called the "average direct effect" (ADE). 
3. The Total Effect, with is ACME + ADE. This is the total effect of the mediator and treatment.

One way to perform a mediation analysis is with the **lavaan** package. ("lavaan" is short for "latent variable analysis".) This is a whole different beast, but the basic idea is you specify a structural equation model as lines of text using lavaan model syntax. Then you fit the model using the `sem()` function. In the formula below, "a", "b", "ADE", and "ACME" are simply labels for the coefficients being estimated. The latter two are what we care about. The `:=` notation says that the ACME is a _latent variable_ that we do not directly observe. So we estimate by multiplying the coefficients "a" and "b".

Here is what the grad student did:

```{r message=FALSE}
library(lavaan)
m <- '
  M ~ a*grp
  Y ~ b*M
  Y ~ ADE*grp
  ACME := a*b 
'
fitm <- sem(m, data=d)
summary(fitm)
```

Now to their question: why is the estimated ADE of 0.110 _so much smaller_ than the estimated effect size of 0.396 from the t-test? Isn't the Direct Effect the same as the effect size returned from the t-test?

As it turns out, no, they are not the same things. The average direct effect (ADE) is the effect of treatment _controlling for mediation_. (I find the terminology of "direct effect" very confusing. especially when the effect is being adjusted for another variable!) The difference in means from a t-test is the effect of treatment _not controlling_ for anything else. It is called the "Total Effect" in mediation analysis. 

If we want to use the lavaan model to estimate the Total Effect (ie, diff in means from t-test), we sum ADE and ACME. So I modified their R code as follows.

```{r}
m <- '
  M ~ a*grp
  Y ~ b*M
  Y ~ ADE*grp
  ACME := a*b 
  Total := ADE + ACME  ## t-test effect size, difference in means
'
fitm <- sem(m, data=d)
summary(fitm)
```

Notice the last line under "Defined Parameters" shows Total as 0.396, which is the same as what we got from the t-test:

```{r}
t.test(Y ~ grp, data = d)$estimate |> 
  diff() |> 
  as.numeric() |>
  round(3)
```

That was basically the consultation. Me figuring out the distinction between Average Direct Effect and Total Effect!

For what it's worth, here's how I simulated the data for this example. First I simulate `M` using `grp` and some noise from a Normal distribution. Then I use both `M` and `grp` to simulate `Y`, one again with some noise. So in the simulation we see that `grp` affected `M`, and then `M` and `grp` affected `Y`. We might call this "partial mediation" since `grp` affects both `M` and `Y`.


```{r eval=FALSE}
n <- 400
set.seed(999)
grp <- sample(0:1, size = n, replace = TRUE) |> factor()
M <- 1.5 + 0.6*(grp=="1") + rnorm(n, sd = 0.1)
Y <- 1.9 + 0.1*(grp=="1") + 0.5*M + rnorm(n, sd = 0.1)
d <- data.frame(grp = grp, Y = Y, M = M)
```

- The TRUE ACME/indirect effect/mediation effect is 0.6 * 0.5 = 0.3. The lavaan model estimated it as 0.286.
- The TRUE ADE = 0.1 (in code where we create `Y`). The lavaan model estimated it as 0.110.
- The TRUE Total Effect (ACME + ADE) = 0.3 + 0.1 = 0.4. The lavaan model estimated it as 0.396.

Notice if we divide ACME by Total Effect we get 0.3/0.4 = 0.75. We might say the proportion of the effect mediated is 0.75. 

The `mediate()` function in the mediation package formally estimates this in addition to ACME, ADE, and Total Effect. Its estimates are slightly different from lavaan's because it uses something called Quasi-Bayesian Approximation which involves simulation. To use the `mediate()` function we first have to fit a model for the mediator and then fit a model for the dependent variable. Below it estimates the proportion of the effect mediated to be about 0.73.

```{r message=FALSE}
library(mediation)
b <- lm(M ~ grp, data = d)
c <- lm(Y ~ grp + M, data = d)
mout <- mediate(b, c, treat = "grp", mediator = "M")
summary(mout)
```

I hope you found this helpful. 
