---
title: "Multilevel Quantile Regression"
author: "Clay Ford"
format:
  html:
    embed-resources: true
---

## Generate data

```{r}
#| code-fold: true
#| code-summary: "Show the code"
n <- 12
id <- 1:n
ln <- 1:2
roi <- 1:3
rep <- 1:50
d <- expand.grid(rep = rep, roi = roi, ln = ln, id = id)
d$trt <- ifelse(d$id < 7, 1, 0)

d$id_ln <- as.numeric(interaction(d$id, d$ln))
d$id_ln_roi <- as.numeric(interaction(d$id, d$ln, d$roi))
set.seed(1)
e <- rnorm(n = nrow(d), sd = ifelse(d$trt == 0,0.5,2*0.5))
e_id <- rnorm(n = n, sd = 0.8)
e_id_ln <- rnorm(n = n*2, sd = 0.5)
e_id_ln_roi <- rnorm(n = n*2*3, sd = 0.7)
d$y <- 4 + 0.4*(d$trt == 1) + 
  e_id[d$id] + 
  e_id_ln[d$id_ln] +
  e_id_ln_roi[d$id_ln_roi] +
  e
d$id_ln <- NULL
d$rep <- NULL
d$trt <- factor(d$trt)
d$roi <- factor(d$roi)
d <- d[,c("id", "ln", "roi", "trt", "y", "id_ln_roi")]
```

## Inspect and visualize data

- id = subject id
- ln = lymph node
- roi = region of interest
- trt = treatment indicator (0 = control/1 = treated)
- y = dependent variable
- id_ln_roi = grouping indicator (id/ln/roi groups)

```{r}
head(d)
```

Plot of y versus treatment group, colored by roi.

```{r}
library(ggplot2)
ggplot(d) +
  aes(x = trt, y = y, color = roi) +
  geom_jitter(width = 0.2, height = 0)
```

Summary of y grouped by trt.

```{r}
tapply(d$y, d$trt, summary)
```


## Multilevel Quantile Model

Model the 0.75 quantile (ie, 75th percentile) as a function of treatment, with a random intercept for the id/ln/roi group. To model a different quantile, change the `tau` argument. 

```{r}
library(lqmm)
fit.lqmm <- lqmm(y ~ trt, random = ~ 1 , 
                 group = id_ln_roi,	
                 data = d, tau = 0.75)
summary(fit.lqmm)
```

We estimate the 75th quantile of treatment is about 1.06 higher than the 75th quantile of the control group.