---
title: "multinomial logistic regression example"
author: "Clay Ford"
date: "2026-04-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load data

Use data from the carData package. Canadian Women's Labour-Force Participation. The Womenlf data frame has 263 rows and 4 columns. The data are from a 1977 survey of the Canadian population.

```{r message=FALSE}
library(carData)
data("Womenlf")

# Labour-Force Participation
summary(Womenlf$partic)
```

## model

Model partic as a function of hincome (husband's income) and children (presence of children in household).

```{r message=FALSE}
library(nnet)
m <- multinom(partic ~ hincome + children, data = Womenlf, 
              Hess = TRUE)
```

Notice the fitted values are probabilities that sum to 1. The first subject would be classified as "not.work" since that category has the highest probability.

```{r}
head(fitted(m))
```

## using a matrix on lhs

Assume that instead of selecting a category, the subjects had allocated percentages to the categories, similar to what Jim describes. Create a left hand side matrix.

```{r}
Y <- round(fitted(m) * 100)
head(Y)
```

Now model the matrix as a function of hincome and children.

```{r}
m2 <- multinom(Y ~ hincome + children, 
               data = Womenlf,
               Hess = TRUE)
```

The fitted values are still probabilities that sum to 1.

```{r}
head(fitted(m2))
```


## visualize model

Probability of "preferring" fulltime decreases as husband's income increases. Likewise, probability of "preferring" not.work increases as husband's income increases

```{r message=FALSE}
library(ggeffects)
plot(ggpredict(m2, terms = "hincome"))
```

