---
title: "Robustness of t-test"
author: "Clay Ford"
date: "1/7/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's simulate data similar to what you sent me via email on 2022 Jan 7.

```{r}
n <- 1800
# the following are based on eyeballing the plot you sent me
probs <- c(25, 50, 75, 200, 325, 225)/
  sum(c(25, 50, 75, 200, 325, 225))

ctrl <- sample(1:6, size = n/2, replace = TRUE, 
               prob = probs)
intv <- sample(1:6, size = n/2, replace = TRUE, 
               prob = probs)
d <- data.frame(score = c(ctrl, intv), 
                group = rep(c("control", "intervention"), each = n/2))
```

Create plot

```{r}
library(ggplot2)
ggplot(d) +
  aes(x = score, fill = group) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Paired") +
  scale_x_continuous(breaks = 1:6)
```

Run Welch t-test

```{r}
t.test(score ~ group, data = d)
```

Ideally this should FAIL TO REJECT the null of no difference between means because I simulated the data for each group from the same distribution. The p-value should be much larger than 0.05.

Now both distributions are distinctly non normal as seen in the plot above. However the t-test comparing the means between these two distributions is _robust to this assumption_. That means it still works. It should falsely reject the null of no difference only about 5% of the time. 

Below we create a function that simulates two data sets from the same non-normal distribution, runs a Welch t-test, and returns TRUE if the p-value is less than 0.05, and FALSE otherwise. 

```{r}
ttest_function <- function(n = 1800){
  ctrl <- sample(1:6, size = n/2, replace = TRUE, 
               prob = probs)
  intv <- sample(1:6, size = n/2, replace = TRUE, 
               prob = probs)
  d <- data.frame(score = c(ctrl, intv), 
                group = rep(c("control", "intervention"), each = n/2))
  t.test(score ~ group, data = d)$p.value < 0.05
}
  
```

Now we replicate the function 2000 times and calculate the proportion of times it rejects the null of no difference (ie, Type 1 error, rejecting the null when it's really true.). 

```{r}
tests <- replicate(n = 2000, ttest_function(n = 1800))
mean(tests)
```

We see that it performs well and that the Type 1 error is close to 0.05 The t-test in this case is robust to departures from normality.