---
title: "splines in mixed-effect models"
author: "Clay Ford"
date: '2022-04-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## generate data

First generate data for 10 subjects (`g`) with 20 observations each (`n`). The `x` predictor ranges from 1:20. The x2 predictor is `x - 13` for `x > 13`. The `r`, `z0`, and `z1` vectors add noise. `z0` adds subject-specific noise to the intercept, `z1` adds subject-specific noise to the slope, and `r` adds noise to all observations. All noise is drawn from a Normal distribution with mean 0 and some fixed standard deviation, as is assumed by linear mixed-effect models. 

```{r}
n <- 20
g <- 10
x <- rep(1:n, g)
id <- rep(1:g, each = n)
x2 <- ifelse((x - 13) > 0, (x - 13), 0)
set.seed(2)
x3 <- runif(n = n * g)
r <- rnorm(length(x), sd = 0.8)
z0 <- rnorm(g, sd = 0.5)
z1 <- rnorm(g, sd = 0.2)
y <- (10 + z0[id]) + (-4 + z1[id])*x + 05*x3 + (6 + z1[id])*x2 + r 
d <- data.frame(id = factor(id), y, x, x3)
```

## plot the data

Quick plot of the data with subjects color-coded.

```{r}
library(ggplot2)
ggplot(d) +
  aes(x, y, color = id, group = id) +
  geom_point() +
  geom_line()
```

Quick plot of the data with a smooth trend line, which in this case is a loess curve. 

```{r}
ggplot(d) +
  aes(x, y) +
  geom_point() +
  geom_smooth()
```

## fit mixed-effect model

Below we fit a mixed-effect model that says the intercept and slope are conditional on subject: `(x|id)`. We also specify that `x` be modeled a cubic spline using the `ns` function from the splines package. `df=3` says we think the trajectory could change direction three times. 

```{r message=FALSE}
library(lme4)
library(splines)
m1 <- lmer(y ~ ns(x, df = 3) + x3 + (x|id), data = d)
```

The summary output is all but impossible to interpret.

```{r}
summary(m1, corr = FALSE)
```

So we can plot the smooth trend line and visually interpret. 

```{r}
library(ggeffects)
eff <- ggpredict(m1, "x[all]")
plot(eff, add.data = TRUE)
```


## piecewise spline

Here we fit something closer to the "correct" model. We use the **segmented** package to find the breakpoint. The **segmented** package does not work with lme4, but does (sort of) work with nlme, an older but still very powerful mixed-effect modeling package. Another quirk is that the variable with the breakpoint cannot have associated random effects, so we have to fit a random-intercept model (`~1|id`). First we fit the model using `lme` and then use `segmented.default` to estimate the breakpoint. The `npsi = 1` argument says we think there is one breakpoint.

```{r message=FALSE, warning=FALSE}
library(segmented)
library(nlme)
m2 <- lme(y ~ x, random = ~ 1|id, data = d)
sm1 <- segmented.default(m2, seg.Z = ~ x, npsi = 1) 
```

Use the `fixef` function to get the estimated slopes for each line. The estimated values are very close to the "true" values we used above to generate the data, -4 and 6. 

```{r}
fixef(sm1)
```

The estimated breakpoint is extracted with the `confint.segmented` function. The true value of 13 falls well within the 95% CI.

```{r}
confint.segmented(sm1, "x", .coef=fixef(sm1))
```

We might report this as follows:

$$y = 10.03 + -3.95x \hspace{1cm} x \le 13.03$$

$$y = 10.03 + -3.95x + 6.05(x - 13.03)\hspace{1cm} x > 13.03$$


Finally we can get a basic plot with a confidence ribbon.

```{r}
plot.segmented(sm1, "x", .coef=fixef(sm1), conf.level=.95)
```

There's probably a way to extract the values and make a nicer looking plot with ggplot2, but I'm not pursuing that.

<br><br><br><br>