---
title: "Term plots"
author: "Clay Ford"
date: "2024-06-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Fit a model with four predictors. All are linear. Notice the residuals versus fitted plot. Not great.

```{r}
data(mtcars)
m <- lm(mpg ~ wt + disp + drat + qsec, data = mtcars)
plot(m, which = 1)
```

Use term plots to investigate departures from linearity for each coefficient. Notice how the smooth trend line for the predictor `wt` _bends_ around the slope. This suggests a quadratic term may be more suitable.

```{r}
op <- par(mfrow=c(2,2))
termplot(m, partial.resid=TRUE, smooth=panel.smooth)
par(op)
```

Add quadratic term for wt predictor. Now the fitted versus residuals plot looks better.

```{r}
m2 <- lm(mpg ~ poly(wt,2) + disp + drat + qsec, data = mtcars)
plot(m2, which = 1)
```


The term plots also look better.

```{r}
op <- par(mfrow=c(2,2))
termplot(m2, partial.resid=TRUE, smooth=panel.smooth)
par(op)

```

The car package also makes these plots using the `crPlots()` function.

```{r message=FALSE}
library(car)
crPlots(m)
```

