---
title: "tidyLPA notes"
author: "Clay Ford"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

- The {tidyLPA} package provides a wrapper for the {mclust} package.

- Latent Profile Analysis (LPA) is a statistical modeling approach for estimating distinct profiles, or groups, of variables.

- LPA can be used for different dimensions, or for a single dimension over time.

- Works best with numeric data.

- The mclust algorithm does not allow for missing data. 

The following example is from 2024-11-01. Data from grad student in Psych department. This was the code she sent us.

```{r message=FALSE}
library(haven)
library(dplyr)
data <- read_sav("data_subset.sav")

# Removing missing cases; The mclust algorithm does not allow for missing data.
data_complete <- data %>%
  filter(!is.na(w2cope) & !is.na(w3cope) & !is.na(w4cope) & !is.na(w5cope) &
         !is.na(w2DISCRIM) & !is.na(w3DISCRIM) & !is.na(w4DISCRIM) & !is.na(w5DISCRIM),
         !is.na(w2GOAL) & !is.na(w3GOAL) & !is.na(w4GOAL) & !is.na(w5GOAL))
```

## Run LPA

Now run a LPA for "coping" measured at four time points. The key {tidyLPA} function is `estimate_profiles()`. 

- The `n_profiles = 2:4` argument says to try 2, 3, and 4 groups, or profiles. 
- The `models = 1` argument says to fit a model with "Equal variances and covariances fixed to 0". In {mclust}, this is model "EEI" with diagonal distribution, equal volume, and equal shape. See Table 3 in [Scrucca, et al 2016](https://journal.r-project.org/articles/RJ-2016-021/).

```{r}
library(tidyLPA)
# Coping LPA
lpa_model <- data_complete %>%
  select(w2cope, w3cope, w4cope, w5cope) %>%
  estimate_profiles(n_profiles = 2:4, models = 1)
```

The output lists the results of three models.

- The BLRT_p column is the result of bootstrap LRT tests. The final p-value of 0.39 suggests the model with 4 classes (or profiles) is no better than the model with 3 classes. The second p-value of 0.01 says the model with 3 classes (or profiles) is preferred to a model with 2 classes.
- The BIC column also selects the model with 3 classes (or profiles)

```{r}
lpa_model
```

To get the same result with {mclust} we use the `Mclust()` function with `G = 2:4` and `modelNames = "EEI"`.

```{r message=FALSE}
library(mclust)
mclust_model <- Mclust(data_complete[,c("w2cope", "w3cope", "w4cope", "w5cope")], 
              G = 2:4, modelNames = "EEI")
```

Calling `summary()` on the model object only returns the best fitting model, the model with 3 classes (or profiles, or "components"). Notice it also assigns observations to the classes, reported in the "Clustering table".

```{r}
summary(mclust_model)
```

## Get estimates

{mclust} estimates means and variances of ellipsoids. Again, see Table 3 in [Scrucca, et al 2016](https://journal.r-project.org/articles/RJ-2016-021/). {tidyLPA} is simply a wrapper for  {mclust}. 

To see the means and variances of a {tidyLPA} object use the `get_estimates()` function. Notice it's returned as a nice data frame.

```{r}
get_estimates(lpa_model)
```

To see the means and variances of a {mclust} object call `summary()` with `parameters = TRUE`. 

```{r}
summary(mclust_model, parameters = TRUE)
```

Notice the "mixing probabilities". These are probabilities for class assignment. The minimum and maximum probabilities are reported in the {tidyLPA} output above as "n_min" and "n_max".

## Plot model

The {tidyLPA} package provides the `plot_profiles()` function, which make a plot for all models that were fit.

```{r}
lpa_model |> 
  plot_profiles(rawdata = FALSE)
```

{mclust} provides a classification plot showing pairwise scatterplots with the classification superimposed. You can see the ellipsoids are all the same size (ie, "Equal variances and covariances fixed to 0")

```{r}
plot(mclust_model, what = "classification")
```

Since this data has a time component, we might want to see how these 3 profiles change over time by plotting the mean versus time. The following code is courtesy of Lauren Brideau:

```{r}
cop_mean <- get_estimates(lpa_model) |> 
  mutate(Class = as.factor(Class),
         Wave = stringr::str_extract(Parameter, "w\\d+")) %>% 
    filter(Category == "Means") 

library(ggplot2)
cop_mean %>% 
  filter(Classes == 3) %>%  
  ggplot(., aes(x = Wave, y = Estimate, group = Class)) +
  geom_line() +
  facet_wrap(vars(Class), labeller = "label_both") +
  theme_bw()
```

We could also add standard error ribbons.

```{r}
cop_mean %>% 
  filter(Classes == 3) %>%  
  ggplot(., aes(x = Wave, y = Estimate, group = Class)) +
  geom_line() +
  geom_ribbon(mapping = aes(ymin = Estimate - se, ymax = Estimate + se), alpha = 1/4) +
  facet_wrap(vars(Class), labeller = "label_both") +
  theme_bw()
```

Use color and combine into one plot.

```{r}
cop_mean %>% 
  filter(Classes == 3) %>%  
  ggplot(., aes(x = Wave, y = Estimate, group = Class, 
                fill = Class, color = Class)) +
  geom_line() +
  geom_ribbon(mapping = aes(ymin = Estimate - se, ymax = Estimate + se), alpha = 1/4) +
  theme_minimal()

```

The grad student who contacted us for help was examining two dimensions: "coping" and "discrimination". She wanted to combine the profiles into two plots. Again, credit to Lauren for the data wrangling and plotting code.

```{r message=FALSE}
# fit the model for discrimination
lpa_discrim <- data_complete %>%
  select(w2DISCRIM, w3DISCRIM, w4DISCRIM, w5DISCRIM) %>%
  estimate_profiles(n_profiles = 2:4, models = 1)
# get means
cop_mean <- get_estimates(lpa_model) %>% 
  mutate(type = as.character("coping"))
dis_mean <- get_estimates(lpa_discrim) %>% 
  mutate(type = as.character("discrimination")) 
#Join data and clean 
cop.dis_mean <- bind_rows(cop_mean, dis_mean) %>% 
  mutate(Parameter = as.factor(Parameter),
         Class = as.factor(Class),
         Wave = stringr::str_extract(Parameter, "w\\d+")) %>% 
    #filter to just the mean values  
  filter(Category == "Means") 
# create the plot
cop.dis_mean %>% filter(Classes == 3) %>%  
  ggplot(., aes(x = Wave, y = Estimate, color = type)) +
  geom_point(aes(shape = Class)) +
  facet_wrap(vars(Class)) +
  geom_line(aes(group = interaction(type, Class))) +
  theme_bw()
```

## Miscellaneous

The documentation for {tidyLPA} states there are only 4 models available when using {mclust}:

- Equal variances and covariances fixed to 0
- Varying variances and covariances fixed to 0
- Equal variances and equal covariances
- Varying variances and varying covariances

These correspond to the following {mclust} models:

- "EEI"
- "VVI"
- "EEE"
- "VVV"

The {mclust} package can fit 14 different models. It's not clear to me why {tidyLPA} is only limited to four of the models. 

The `mclustBIC()` function allows us try all models and group sizes (or profiles) ranging from 1 - 9. For example:

```{r message=FALSE}
out_bic <- mclustBIC(data_complete[,c("w2cope", "w3cope", "w4cope", "w5cope")])
summary(out_bic)
```

The "best" model is "VEI" with 6 profiles, a model that is not available in {tidyLPA}. We can then fit the model using `Mclust()`.

```{r}
mod <- Mclust(data_complete[,c("w2cope", "w3cope", "w4cope", "w5cope")], x = out_bic)
summary(mod)
```

However, the classification plot has me wondering if maybe 3 or 4 groups would be better.

```{r}
plot(mod, what = "classification")
```

See also this [blog post](https://darrendahly.github.io/post/cluster/) on the dangers of falling in love with clustering.

## References

- Scrucca, et al., "mclust 5: Clustering, Classification and Density Estimation Using Gaussian Finite Mixture Models", The R Journal, 2016. <https://journal.r-project.org/articles/RJ-2016-021/>
- Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Van Lissa, C. J., & Schmidt, J.
  A. (2018). tidyLPA: An R Package to Easily Carry Out Latent Profile Analysis (LPA)
  Using Open-Source or Commercial Software. Journal of Open Source Software, 3(30),
  978, https://doi.org/10.21105/joss.00978
- Scrucca L, Fraley C, Murphy TB, Raftery AE (2023). _Model-Based Clustering,
  Classification, and Density Estimation Using mclust in R_. Chapman and Hall/CRC.
  ISBN 978-1032234953, doi:10.1201/9781003277965
  <https://doi.org/10.1201/9781003277965>, <https://mclust-org.github.io/book/>.
  
## Session information

```{r}
sess <- sessionInfo()
sess$R.version$version.string
sess$platform
packageVersion("tidyLPA")
packageVersion("mclust")
packageVersion("dplyr")
packageVersion("haven")
```

