---
title: "Tukey HSD"
author: "Clay Ford"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simulate data

Simulate three groups of data by sampling from three normal distributions with the same standard deviation (1) but different means. The `set.seed(123)` line makes this data reproducible.

```{r}
set.seed(123)
a <- rnorm(n = 20, mean = 3, sd = 1)
b <- rnorm(n = 20, mean = 3.5, sd = 1)
c <- rnorm(n = 20, mean = 4, sd = 1)
d <- data.frame(grp = rep(c("a", "b", "c"), each = 20),
                y = c(a, b, c))
```

Calculate the means for each group:

```{r}
means <- aggregate(y ~ grp, data = d, mean)
means
```

Calculate difference in means:

```{r}
ba <- means$y[means$grp == "b"] - means$y[means$grp == "a"]
ca <- means$y[means$grp == "c"] - means$y[means$grp == "a"]
cb <- means$y[means$grp == "c"] - means$y[means$grp == "b"]
results <- data.frame(comparison = c("b-a", "c-a", "c-b"), 
           diff = c(ba, ca, cb))
results
```


## T-tests

Run t-tests for all three comparisons:

```{r}
ba <- t.test(y ~ grp, data = d, subset = grp != "c")
ca <- t.test(y ~ grp, data = d, subset = grp != "b")
cb <- t.test(y ~ grp, data = d, subset = grp != "a")
```

Put p-values in a table:

```{r}
results$pvalue <- c(ba$p.value, ca$p.value, cb$p.value)
results
```

Notice two are "significant" at the 0.05 level.

## ANOVA and Tukey's HSD

Now analyzing using a one-way ANOVA and conduct pairwise comparisons using Tukey's HSD. Notice all p-values are higher than their t-test counterparts and only one is "significant" at 0.05 level.

```{r}
m <- aov(y ~ grp, data = d)
TukeyHSD(m)
```

