---
title: "Using polycor::hetcor"
author: "Clay Ford"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Demonstration of using the `hetcor()` function in the {polycor} package using simulated data. From the help page: "hetcor computes a heterogenous correlation matrix, consisting of Pearson product-moment correlations between numeric variables, polyserial correlations between numeric and ordinal variables, and polychoric correlations between ordinal variables."

Load some packages.

```{r message=FALSE}
library(lavaan) # for getCov() function
library(mvtnorm) # for generating multivariate data
library(scales) # for rescale() function
library(polycor) # for hetcor() function
```

First create a covariance matrix. The {lavaan} package has a nice function for this. First create the lower portion of the matrix as a text string with line breaks, then use the `getCov()` function. I'm creating a correlation matrix to make life easier.

```{r}
lower <- '
1
0.5 1
0.6 -0.3 1
0.8 0.2 0.7 1'
S <- getCov(lower)
S
```

(See also the examples on the help page for `hetcor()` for an alternative way to generate a random covariance matrix using matrix algebra.)

Next generate some multivariate normal data using our covariance matrix and convert to data frame

```{r}
set.seed(123)
d <- as.data.frame(rmvnorm(n = 1000 * 4, mean = rep(1, 4), sigma = S))
```

Create ordered likert-type variables that range from 1 - 5. The `rescale()` function is handy for this.

```{r}
d$V1 <- factor(round(rescale(d$V1, to = c(1,5))))
d$V2 <- factor(round(rescale(d$V2, to = c(1,5))))
```

Glance at the data.

```{r}
head(d)
```

Now we're ready to use the `hetcor()` function to generate a correlation matrix for this data. It will automatically calculate either polychoric, polyserial, or pearson correlations depending on data type.

```{r}
cor_d <- hetcor(d)
cor_d
```

If desired we can extract the correlation matrix from the `hetcor` object.

```{r}
cor_d$correlations
```

This can be useful creating visualizations such as the kind available with the {corrplot} package.

```{r message=FALSE}
library(corrplot)
corrplot(cor_d$correlations, type = 'upper', diag = FALSE)
```

