---
title: "VarCov Matrices for Random Effects"
author: "Clay Ford"
format: 
  html:
    embed-resources: true
    
---

In this note I show how to suppress estimation of covariance random effects when estimating random effects for a categorical predictor.

<hr>

The following data come from Ch 5 of _Linear Mixed Models_ by West et al.

An experiment that examined nucleotide activation in 5 different rats. Each
rat was subjected to two different treatments (basal vs carbachol) and then
nucleotide activation was measured in 3 different brain regions. Therefore
each rat has 6 different measures.

```{r message=FALSE}
library(lme4)
library(afex)
library(nlme)
ratbrain <- read.csv("ratbrain.csv")
head(ratbrain, n = 8)
interaction.plot(x.factor = ratbrain$region,
                 trace.factor = ratbrain$treatment, 
                 response = ratbrain$activate)
```

Model `activate` as a function of region and treatment (and their interaction) with random effects on the intercept and region. Notice the singular fit message, This is due to the ill-conditioned random-effects covariance matrix, which suggests the model may be over-parameterized. Specifically we may not need to model covariance between the levels of the region random effects. The summary shows them all as -1 or 1.

```{r}
m1 <- lmer(activate ~ region * treatment + (region | animal), 
            data = ratbrain)
summary(m1, corr = F)
```

How can we fit a model without the covariance between random effects? Unfortunately, we can't do that in {lme4}. However we can do so with the {afex} or {nlme} packages.

The {afex} package provides the `lmer_alt()` function that allows the use of two bars between region and animal (`region || animal`) to suppress estimation of random effect covariances (i.e., they are held constant at 0). However we still get warning about singularities. 

```{r}
m2 <- lmer_alt(activate ~ region * treatment + (region || animal), 
            data = ratbrain)
summary(m2, corr = F)
```

In the {nlme} package we can use the `pdDiag()` to specify we want to fit a _diagonal_ random effects matrix with covariances fixed at 0. Notice this requires a named list object, where the name is the grouping factor. This converges with no warning messages. This may be due to `lme()` using a different optimizer. It uses "BFGS" while `lmer()` uses "nloptwrap". 

```{r}
library(nlme)
m3 <- lme(activate ~ region * treatment, 
          random = list(animal = pdDiag(~ region)), 
          data = ratbrain)
summary(m3)
```

### References

West, B., Welch, K., & Galecki, A. (2015) *Linear Mixed Models*. Chapman and Hall/CRC.
